EGGs are adhesive!\\ An algebraic account of equality saturation

### 2025-08-19 jlamp/jlamp.tex
fine 3.1

\def\R{\mathsf{R}}
\def\B{\textbf {\textup{B}}}
\def\C{\textbf {\textup{C}}}
\def\D{\textbf {\textup{D}}}
\def\X{\textbf {\textup{X}}}
\def\Y{\textbf {\textup{Y}}}
\def\E{\textbf {\textup{E}}}
\def\T{\textbf {\textup{1}}}
\def\A{\textbf {\textup{A}}}
\def\M{\mathcal{M}}
%categorie varie
\newcommand{\catname}[1]{\textbf{\textup{#1}}}
\newcommand{\lab}{\catname{LHyp}}
\newcommand{\lmo}{\catname{LMH}}
\newcommand{\hyp}{\catname{Hyp}}
\newcommand{\shyp}{\catname{SepHyp}}
\newcommand{\hyps}{\catname{Hyp}_{\Sigma}}
\newcommand{\EqHyp}{\catname{EqHyp}} %equivalence hypergraphs
\newcommand{\EqHyps}{\catname{EqHyp}_{\Sigma}}
\newcommand{\EqTG}{\catname{EqTG}}
\newcommand{\EqTGs}{\catname{EqTG}_{\Sigma}}
\newcommand{\GEqTGs}{\catname{GEqTG}_{\Sigma}}
\newcommand{\GEqATGs}{\catname{GEqATG}_{\Sigma}}
\newcommand{\gr}{\textbf{\textup{Graph}}}
\newcommand{\dgr}{\catname{SGraph}}
\newcommand{\dg}{\catname{DAG}}
\newcommand{\rt}{\mathsf{dcl_s}}
\newcommand{\rta}{\mathsf{dcl}}
\newcommand{\rtd}{\mathsf{dcl_{d}}}
\newcommand{\slice}[2]{(\catname{#1}\downarrow{#2})}
\newcommand{\tg}[0]{\catname{TG}_{\Sigma}}
\newcommand{\teg}[0]{\catname{TeGr}_{\Sigma}}
\newcommand{\sv}[0]{\mathsf{Sieves}}
\newcommand{\mono}[1]{\mathsf{Mon}(\catname{#1})}
\newcommand{\mo}[1]{{#1}_\mathsf{Mon}}
\newcommand{\pro}{\mathsf{prod}}
\newcommand{\spro}{\mathsf{ps}}
\newcommand{\prol}{\mathsf{lprod}}
\newcommand{\pred}[1]{{\downarrow}#1}
\newcommand{\colim}[0]{\mathrm{colim}}
\newcommand{\cod}{\mathsf{cod}}
\renewcommand{\sp}{\mathsf{sp}}
\newcommand{\is}[1]{{#1}_{is}}
\renewcommand{\sup}{\mathsf{sup}}
\newcommand{\cow}[1]{\mathsf{cwd}({#1})}
\renewcommand{\inf}{\mathsf{inf}}
\newcommand{\dom}{\mathsf{dom}}
\newcommand{\dwnarrow}{\downarrow \hspace{-2pt}}
\newcommand{\Dwnarrow}{\Downarrow \hspace{-2pt}}
\newcommand{\egg}{\mathsf{e}\text{-}\catname{EqHyp}}

@@

\title{EGGs are adhesive!}

@@

\begin{abstract}
The use of rewriting-based visual formalisms is on the rise. 
%
In the formal methods community, this is due also to the introduction of adhesive
categories, where most properties of classical approaches to graph transformation, 
such as those on parallelism and confluence, can be rephrased and proved in a general and 
uniform way.
%
E-graphs (EGGs) are a formalism for program optimisation 
via an efficient implementation of equality saturation. 
In short, EGGs can be  defined as (acyclic) term graphs with an additional notion of 
equivalence on nodes that is closed under the operators of the signature.
Instead of replacing the components of a program, the optimisation step 
is performed by adding new components and linking them to 
the existing ones via an equivalence relation, until an optimal program is reached.
%
This work describes EGGs via adhesive categories. 
Besides the benefits in itself of a formal presentation, which renders the 
properties of the data structure precise, the description of the addition of equivalent 
program components using standard graph transformation tools offers the advantages 
of the adhesive framework in modelling, for example, concurrent updates.
%
%Besides 
%the benefits \emph{per se} of a formal presentation, making precise the properties of 
%the data structure, describing the addition of equivalent program components 
%via the standard tools of graph transformation offers the advantages given by the 
%adhesive framework in modelling e.g. concurrent updates.
\end{abstract}

@@

\section{Introduction}
The introduction of \emph{adhesive categories} marked a watershed moment for the algebraic approaches 
to the rewriting of graph-like structures~\cite{lack2005adhesive,ehrig2006fundamentals}.
%
Until then, key results of the approaches on e.g. parallelism and confluence had to be proven 
over and over again for each different formalism at hand, %let it be either xxx or xx, 
despite the obvious similarity of the procedure.
%
Adhesive categories provides such a disparate set of formalisms with a common abstract framework 
where many of these general results can be recast and uniformly proved once and for all.

\vspace{.1cm}
\noindent
\begin{minipage}[l]{.78\linewidth}
		\setlength{\parindent}{1.5em}
		In short, following the double-pushout (DPO) approach
to graph transformation~\cite{CorradiniMREHL97,ehrig2006fundamentals}, 
%~\cite{ehrig1999handbook,ehrig2006fundamentals,corradini1997algebraic}, 
a rule is given by two arrows $l: K \to L$ and $r: K \rightarrow R$
and its application requires a match $m: L \to G$: the rewriting step from $G$
to $H$ is then given by the diagram aside, whose halves are pushouts.
  \end{minipage}%
    \hfill 
  \begin{minipage}[r]{.20\linewidth }
    \xymatrix@C=.5cm@R=.5cm{
      L \ar[d]_{m}
      & K \ar[r]^r \ar[l]_{l} \ar[d]_{k} & R \ar [d]^h \\
      G & C \ar[r]_{g} \ar[l]^{f}                    & H
    }
%    \xymatrix@C=.5cm@R=.5cm{
%      L \ar[d]_{m} \ar@{}[dr]|{(1)}
%      & I \ar[r]^r \ar@{>->}[l]_{l} \ar[d] \ar@{}[dr]|{(2)} & R \ar [d] \\
%      G & C \ar@{->}[r] \ar@{>->}[l]                    & H\\
 %     & J \ar@{->}[ul] \ar@{->}[u]|{k} \ar@{->}[ur]
 %   }
  \end{minipage}

Thus, $L$ and $R$ are the left- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet that are not affected by the rule itself.
%
Should a category be adhesive, and the arrows of the rules monomorphisms, the presence of a match ensures that
if the pushout complement $C \to G$ exists then it is unique, hence a rewriting step can be deterministically performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{azzi2019essence,heindel2009category} extends the core framework, ensuring that if the arrows 
of the rules are in a suitable class $\mathcal{M}$ of monomorphisms
%and the match is in $\mathcal{N}$, 
then the benefits of adhesivity
%the classical results from the theory of graph transformation 
can be recovered~\cite{ehrig2012,ehrig2014adhesive}. 
If only the left-hand side belongs to $\mathcal{M}$, the theory is still under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness,
% and the immediate advantages, 
%to be adhesive, 
proving that a given category satisfies the conditions 
for being $\mathcal{M}$-adhesive can be a daunting task. For this reason, sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their $\mathcal{M}$-adhesivity has not yet been settled.

@@

EGGs can be concisely defined as term graphs equipped with a notion of equivalence on nodes
that is closed under the operators of a signature~\cite[Section~4.2]{DetlefsNS05}.
%(i.e., graphs whose components are labelled with the operators of a signature)
In the presentation of term graphs via string diagrams~\cite{CastelnovoGM24}, EGGs are (hyper)trees 
whose edges are labelled by operators and with
the possibility of sharing subtrees, with an additional equivalence relation $\equiv$ on nodes that 
is closed under composition. In plain words and using a toy example:
if $a$ and $b$ are two constants such that $a \equiv b$, then $f(a) \equiv f(b)$ for any unary operator $f$.

@@

Building on the criterion developed in~\cite{CastelnovoGM24}, this work proves that both hypergraphs with equivalence
and EGGs form an 
$\mathcal{M}$-adhesive category for a suitable choice of $\mathcal{M}$.
The advantages from this characterisation are two-fold. On the one side, 
we put the benefits \emph{per se} of a formal presentation, making precise the properties of the data structure. 
On the other side, describing the optimisation steps via the DPO approach
offers the tools for modelling their parallel and concurrent execution
and for proving their confluence and termination.

@@

We open this background section by fixing some notation. Given a category $\X$ we do not distinguish notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$. We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of all arrows, monos and regular monos of $\X$, respectively.  Given an object $X$, we  denote by $?_X$ the unique arrow from an initial object into $X$ and by $!_X$ that  unique arrow from $X$ into a terminal one. We will also use the notation $e\colon X\eto Y$ to denote that an arrow $e\colon X\to Y$ is a regular epi.

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property of $\mathcal{M}$-adhesive categories is the \emph{Van Kampen condition}~\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive},
%
and for defining it we need some notions.
%and to define it we introduce some notions.  
Let  $\X$ be a category. A subclass $\mathcal{A}$ of $\mor(\X)$ is said to be:

\noindent
\begin{minipage}[l]{.85\linewidth}
	\begin{itemize}
		\item		\emph{stable under pushouts (pullbacks)} if for every pushout (pullback) square as the one aside, if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
		\item \emph{closed under composition} if $h, k\in \mathcal{A}$ implies $h\circ k\in \mathcal{A}$ whenever $h$ and $k$ are composable;
	\end{itemize}
\end{minipage}\hfill 
\begin{minipage}[r]{.13\linewidth}
\xymatrix{A \ar[r]^{f} \ar[d]_{m}& B \ar[d]^{n} \\ C \ar[r]_{g} & D}
\end{minipage}
\begin{itemize}	\item \emph{closed under decomposition}, or \emph{left-cancellable}, if whenever $g$ and $g\circ f$ belong to $\mathcal{A}$, then $f\in \mathcal{A}$.
\end{itemize}

	\begin{defi}
	Let $\mathcal{A}\subseteq \mor(\X)$ be a class of arrows in a category $\X$ and consider the cube below on the right.

	\begin{minipage}[l]{.73\linewidth}
		We say that the bottom square is an \emph{$\mathcal{A}$-Van Kampen square} if:
	\begin{enumerate}
		\item it is a pushout square;
		\item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
		if and only if the front and right faces are pullbacks.
	\end{enumerate}
\end{minipage}\hfill 
\begin{minipage}[r]{.25\linewidth}
	\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D }
\end{minipage}

	Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}. A $\mor(\X)$-Van Kampen square is called  \emph{Van
		Kampen} and a $\mor(\X)$-stable square  \emph{stable}.
\end{defi}

@@

\begin{prop}
	\label{thm:slice-functors}
	Let $\X$ be an (strict) $\mathcal{M}$-adhesive category. Then the following hold
	\begin{enumerate}
		\item if $\Y$ is an (strict) $\mathcal{N}$-adhesive category, $L\colon \Y\to \A$ a functor preserving $\mathcal{N}$-pushouts and $R\colon \X\to \A$ one preserving $\mathcal{M}$-pullbacks, then $\comma{L}{R}$ is (strictly) $\comma{\mathcal{N}}{\mathcal{M}}$-adhesive, where
		\[\comma{\mathcal{N}}{\mathcal{M}}:=\{(h,k) \in \mor(\comma{L}{R}) \mid h\in \mathcal{N}, k\in \mathcal{M}\}\]
		\item for every object $X$
		the categories $\X/X$  and $X/\X$ are, respectively, (strictly) $\mathcal{M}/X$-adhesive and (strictly) $X/\mathcal{M}$-adhesive, where
		\[\mathcal{M}/X:=\{m\in \mor(\X/X) \mid m\in
		\mathcal{M}\} \hspace{25pt} X/\mathcal{M}:=\{m\in \mor(X/\X) \mid m\in \mathcal{M}\}\]
		\item for every small category $\Y$, the category $\X^\Y$ of
		functors $\Y\to \X$ is (strictly) $\mathcal{M}^{\Y}$-adhesive, where
		$\mathcal{M}^{\Y}:=\{\eta \in \mor(\X^\Y) \mid \eta_Y \in
		\mathcal{M} \text{ for every } Y\in \Y\}$;
		\item if $\Y$ is a full subcategory of $\X$ closed under pullbacks and $\mathcal{M}$-pushouts, then $\Y$ is (strictly) $\mathcal{N}$-adhesive for every class of arrows $\mathcal{N}$ of $\Y$ contained in $\mathcal{M}$ that is stable under pullbacks and pushouts, contains all isos, and is closed under composition and decomposition.
	\end{enumerate} 
\end{prop}

@@

\begin{exa}
	\label{ex:adhesive}
	$\cat{Set}$ is adhesive, and, more generally, every topos is
	adhesive~\cite{lack2006toposes}. By the closure properties above, every presheaf $[\cat{X},\cat{Set}]$ is adhesive, thus the category
	$\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is adhesive
	where $E \rightrightarrows {V}$ is the two objects category with two
	morphisms $s,t \colon{E} \to {V}$. Similarly, various
	categories of hypergraphs can be shown to be adhesive, such as term
	graphs and hierarchical graphs~\cite{CastelnovoGM24}. Note that the category $\cat{sGraphs}$ of simple graphs, 
	i.e.~graphs without parallel edges, is
	$\reg{(\cat{sGraphs})}$-adhesive~\cite{BehrHK23} but not
	quasiadhesive.
\end{exa}

@@

We can state some useful properties of $\mathcal{M}$-adhesive category
(see, for instance, \cite[Thm.~4.26]{ehrig2006fundamentals} or \cite[Fact 2.6]{azzi2019essence}). A proof is provided in \Cref{regmono-proof}.
\begin{restatable}{prop}{regmono}\label{prop:regmono}
	Let $\X$ be an $\mathcal{M}$-adhesive category. Then the following hold
	\begin{enumerate}
		\item every $\mathcal{M}$-pushout square is also a pullback;
		\item every arrow in $\mathcal{M}$ is a regular mono.
	\end{enumerate}
\end{restatable}
%  [Proof in ]

@@

\begin{restatable}{prop}{epic}\label{prop:reg_epi_coeq}
    Let $e\colon X \eto Y$ be a regular epi in a category $\X$ with a kernel pair $(K_e, \pi^1_e, \pi^2_e)$. Then, $e$ is the coequalizer of $\pi^1_e$ and $\pi^2_e$.
\end{restatable}
%[Proof in \Cref{epic-proof}]

@@

		Moreover, the following hold
		\begin{enumerate}
			\item if $h$ is a mono then $k_h$ is a mono;
			\item if the leftmost square is a pullback then the central two are pullbacks;
			\item if $h$ is mono and the leftmost square is a pullback then the rightmost is a pullback.
		\end{enumerate} 
\end{restatable}
%[Proof in \Cref{kpp-proof}]

@@

\noindent
\begin{minipage}[l]{.6\linewidth}
\begin{restatable}{prop}{mpo}\label{lem:mpo}
	Let $\X$ be a strict $\mathcal{M}$-adhesive category with all pullbacks, and suppose that in the cube aside the top face is an $\mathcal{M}$-pushout and all the vertical faces are pullbacks. Then the right square is a pushout.
\end{restatable}
\end{minipage}\hfill 
\begin{minipage}[r]{.38\linewidth}\xymatrix@C=10pt@R=2pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]_(.3){f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]_{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]_(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]_{m} && B \ar[dl]^{n}  & K_{c} \ar[rr]_{k_{g'}}&& K_d\\C \ar[rr]^{g} & & D }
	\end{minipage}

@@

\begin{cor}\label[corollary]{cor:monhyper}
	A morphism $(h,k)$ is a mono in $\hyp$ if and only if both its components are injective functions.
\end{cor}

@@

\subsubsection{Some  subcategories of $\hyp$}

>>>>

\def\R{\mathsf{R}}
\def\B{\textbf {\textup{B}}}
\def\C{\textbf {\textup{C}}}
\def\D{\textbf {\textup{D}}}
\def\X{\textbf {\textup{X}}}
\def\Y{\textbf {\textup{Y}}}
\def\E{\textbf {\textup{E}}}
\def\T{\textbf {\textup{1}}}
\def\A{\textbf {\textup{A}}}
\def\M{\mathcal{M}}
%categorie varie
\newcommand{\catname}[1]{\textbf{\textup{#1}}}
\newcommand{\lab}{\catname{LHyp}}
\newcommand{\lmo}{\catname{LMH}}
\newcommand{\hyp}{\catname{Hyp}}
\newcommand{\phyp}{\catname{pHyp}}
\newcommand{\pshyp}{\catname{pSep}}
\newcommand{\shyp}{\catname{Sep}}
\newcommand{\hyps}{\catname{Hyp}_{\Sigma}}
\newcommand{\EqHyp}{\catname{EqHyp}} %equivalence hypergraphs
\newcommand{\EqHyps}{\catname{EqHyp}_{\Sigma}}
\newcommand{\EqTG}{\catname{EqTG}}
\newcommand{\EqTGs}{\catname{EqTG}_{\Sigma}}
\newcommand{\GEqTGs}{\catname{GEqTG}_{\Sigma}}
\newcommand{\GEqATGs}{\catname{GEqATG}_{\Sigma}}
\newcommand{\gr}{\textbf{\textup{Graph}}}
\newcommand{\dgr}{\catname{SGraph}}
\newcommand{\dg}{\catname{DAG}}
\newcommand{\rt}{\mathsf{dcl_s}}
\newcommand{\rta}{\mathsf{dcl}}
\newcommand{\rtd}{\mathsf{dcl_{d}}}
\newcommand{\slice}[2]{(\catname{#1}\downarrow{#2})}
\newcommand{\tg}[0]{\catname{TG}_{\Sigma}}
\newcommand{\teg}[0]{\catname{TeGr}_{\Sigma}}
\newcommand{\sv}[0]{\mathsf{Sieves}}
\newcommand{\mono}[1]{\mathsf{Mon}(\catname{#1})}
\newcommand{\mo}[1]{{#1}_\mathsf{Mon}}
\newcommand{\pro}{\mathsf{prod}}
\newcommand{\spro}{\mathsf{ps}}
\newcommand{\prol}{\mathsf{lprod}}
\newcommand{\pred}[1]{{\downarrow}#1}
\newcommand{\colim}[0]{\mathrm{colim}}
\newcommand{\cod}{\mathsf{cod}}
\renewcommand{\sp}{\mathsf{sp}}
\newcommand{\is}[1]{{#1}_{is}}
\renewcommand{\sup}{\mathsf{sup}}
\newcommand{\cow}[1]{\mathsf{cwd}({#1})}
\renewcommand{\inf}{\mathsf{inf}}
\newcommand{\dom}{\mathsf{dom}}
\newcommand{\dwnarrow}{\downarrow \hspace{-2pt}}
\newcommand{\Dwnarrow}{\Downarrow \hspace{-2pt}}
\newcommand{\egg}{\mathsf{e}\text{-}\catname{EqHyp}}

@@

\title{EGGs are adhesive!\\ An algebraic account of equality saturation}

@@

\begin{abstract}
The use of rewriting-based visual formalisms is on the rise. 
%
In the formal methods community, this is partly due to the introduction of adhesive
categories, where most properties of classical approaches to graph transformation, 
such as those on parallelism and confluence, can be rephrased and proved in a general and 
uniform way.
%
E-graphs (EGGs) are a formalism for program optimisation 
via an efficient implementation of equality saturation. 
In short, EGGs are (acyclic) term graphs with an additional
equivalence on nodes that is closed under the operators of the signature.
Instead of replacing the components of a program, the optimisation step 
adds new components and links them to 
the existing ones via an equivalence relation, until an optimal program is reached.
%
This work describes EGGs via adhesive categories. 
Besides the benefits in itself of a formal presentation, which renders the 
properties of the data structure precise, the description of the addition of equivalent 
program components using standard graph transformation tools offers the advantages 
of the adhesive framework in modelling, for example, concurrent updates.
%
%Besides 
%the benefits \emph{per se} of a formal presentation, making precise the properties of 
%the data structure, describing the addition of equivalent program components 
%via the standard tools of graph transformation offers the advantages given by the 
%adhesive framework in modelling e.g. concurrent updates.
\end{abstract}

@@

\section{Introduction}
The introduction of \emph{adhesive categories} marked a watershed moment for the algebraic approaches 
to the rewriting of graph-like structures~\cite{lack2005adhesive,ehrig2006fundamentals}.
%
Until then, key results of the approaches on e.g. parallelism and confluence had to be proven 
over and over again for each different formalism at hand, %let it be either xxx or xx, 
despite the obvious similarity of the procedure.
%
Adhesive categories provides such a disparate set of formalisms with a common abstract framework 
where many general results can be recast and uniformly proved once and for all.

%\vspace{.1cm}
\noindent
\begin{minipage}[l]{.78\linewidth}
		\setlength{\parindent}{1.5em}
		In short, following the double-pushout (DPO) approach
to graph transformation~\cite{CorradiniMREHL97,ehrig2006fundamentals}, 
%~\cite{ehrig1999handbook,ehrig2006fundamentals,corradini1997algebraic}, 
a rule is given by two arrows $l: K \to L$ and $r: K \rightarrow R$
and its application requires a match $m: L \to G$: the rewriting step from $G$
to $H$ is then given by the diagram aside, whose halves are pushouts.
  \end{minipage}%
    \hfill 
  \begin{minipage}[r]{.20\linewidth }
    \xymatrix@C=.5cm@R=.5cm{
      L \ar[d]_{m}
      & K \ar[r]^r \ar[l]_{l} \ar[d]_{k} & R \ar [d]^h \\
      G & C \ar[r]_{g} \ar[l]^{f}                    & H
    }
%    \xymatrix@C=.5cm@R=.5cm{
%      L \ar[d]_{m} \ar@{}[dr]|{(1)}
%      & I \ar[r]^r \ar@{>->}[l]_{l} \ar[d] \ar@{}[dr]|{(2)} & R \ar [d] \\
%      G & C \ar@{->}[r] \ar@{>->}[l]                    & H\\
 %     & J \ar@{->}[ul] \ar@{->}[u]|{k} \ar@{->}[ur]
 %   }
  \end{minipage}
  \vspace{.01cm}

Thus, $L$ and $R$ are the left- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet that are not affected by the rule itself.
%
If a category is adhesive, and the arrows of the rules are monomorphisms, the presence of a match ensures that
if the pushout complement $C \to G$ exists then it is unique, hence a rewriting step can be deterministically performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{azzi2019essence,heindel2009category} extends the core framework, ensuring that if the arrows 
of the rules are in a suitable class $\mathcal{M}$ of monomorphisms
%and the match is in $\mathcal{N}$, 
then the benefits of adhesivity
%the classical results from the theory of graph transformation 
can be recovered~\cite{ehrig2012,ehrig2014adhesive}. 
If only the left-hand side belongs to $\mathcal{M}$, the theory is still under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness,
% and the immediate advantages, 
%to be adhesive, 
proving that a category satisfies the conditions 
for being $\mathcal{M}$-adhesive can be a daunting task. For this reason, sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for %the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their $\mathcal{M}$-adhesivity has not yet been settled.

@@

EGGs can be concisely defined as term graphs equipped with an equivalence on nodes
that is closed under the operators of a signature~\cite[Section~4.2]{DetlefsNS05}.
%(i.e., graphs whose components are labelled with the operators of a signature)
In the presentation of term graphs via string diagrams~\cite{CastelnovoGM24}, EGGs are (hyper)trees 
whose edges are labelled by operators and with
the possibility of sharing subtrees, with an equivalence relation $\equiv$ on nodes that 
is closed under composition. In plain words and using a toy example:
if $a$ and $b$ are two constants such that $a \equiv b$, then $f(a) \equiv f(b)$ for any unary operator $f$.

@@

Building on the criterion developed in~\cite{CastelnovoGM24}, this work proves that both hypergraphs with equivalence
and EGGs form an 
$\mathcal{M}$-adhesive category for a suitable choice of $\mathcal{M}$.
The advantages from this characterisation are two-fold. On the one side, 
we put the benefits \emph{per se} of a formal presentation, making precise the properties of the data structure. 
On the other side, describing the optimisation steps via the DPO approach
offers the tools for modelling  parallel and concurrent execution
and for proving  confluence and termination.

@@

We open this background section by fixing notation. Given a category $\X$ we do not distinguish notationally between $\X$ and its class of objects, so
``$X\in \X$'' means that $X$ is an object of $\X$. We let $\mor(\X)$, $\mon(\X)$ and $\reg(\X)$ denote the class of all arrows, monos and regular monos of $\X$, respectively.  Given an object $X$, $?_X$ denotes the unique arrow from an initial object into $X$ and by $!_X$ that  unique arrow from $X$ into a terminal one. We also use the notation $e\colon X\eto Y$ to denote that an arrow $e\colon X\to Y$ is a regular epi.

\subsection{$\mathcal{M}$-adhesivity}\label{subsec:ade}
The key property of $\mathcal{M}$-adhesive categories is the \emph{Van Kampen condition}~\cite{brown1997van,johnstone2007quasitoposes,lack2005adhesive},
%
and for defining it we need some notions.
%and to define it we introduce some notions.  
Let  $\X$ be a category. A subclass $\mathcal{A}$ of $\mor(\X)$ is said to be

\noindent
\begin{minipage}[l]{.85\linewidth}
\begin{itemize}
	\item		\emph{stable under pushouts (pullbacks)} if for every pushout (pullback) square as the one aside, if $m \in \mathcal{A}$ ($n\in \mathcal{A}$) then $n \in \mathcal{A}$ ($m \in \mathcal{A}$);
	\item \emph{closed under composition} if $h, k\in \mathcal{A}$ implies $h\circ k\in \mathcal{A}$ whenever $h$ and $k$ are composable;
	\item \emph{closed under decomposition}, or \emph{left-cancellable}, if whenever $g$ and $g\circ f$ belong to $\mathcal{A}$, then $f\in \mathcal{A}$.
	\end{itemize}
\end{minipage}\hfill 
\begin{minipage}[r]{.13\linewidth}
\xymatrix{A \ar[r]^{f} \ar[d]_{m}& B \ar[d]^{n} \\ C \ar[r]_{g} & D}
\end{minipage}
%\begin{itemize}
%\end{itemize}

	\begin{defi}
	Let $\mathcal{A}\subseteq \mor(\X)$ be a class of arrows in a category $\X$ and consider the cube below. %on the right. 
	We say that the bottom square is an \emph{$\mathcal{A}$-Van Kampen square} if
\vspace{-.5cm}

	\begin{minipage}[l]{.68\linewidth}
	\begin{enumerate}
		\item it is a pushout square;
		\item 	whenever the cube above has pullbacks as back and left faces and the vertical arrows belong to $\mathcal{A}$, then its top face is a pushout 
		if and only if the front and right faces are pullbacks.
	\end{enumerate}
\end{minipage}\hfill 
\begin{minipage}[r]{.3\linewidth}
	\xymatrix@C=10pt@R=6pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar[dl]_{m'} && B' \ar[dd]^{b} \ar[dl]_{n'} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]^{m} && B \ar[dl]^{n} \\C \ar[rr]_{g} & & D }
\end{minipage}
	Pushout squares that enjoy only the ``if'' half of item (2) above are called \emph{$\mathcal{A}$-stable}. A $\mor(\X)$-Van Kampen square is called  \emph{Van
		Kampen} and a $\mor(\X)$-stable square  \emph{stable}.
\end{defi}

@@

\begin{thm}
	\label{thm:slice-functors}
	Let $\X$ be an (strict) $\mathcal{M}$-adhesive category. Then the following facts hold
	\begin{enumerate}
		\item if $\Y$ is an (strict) $\mathcal{N}$-adhesive category, $L\colon \Y\to \A$ a functor preserving $\mathcal{N}$-pushouts and $R\colon \X\to \A$ one preserving $\mathcal{M}$-pullbacks, then $\comma{L}{R}$ is (strictly) $\comma{\mathcal{N}}{\mathcal{M}}$-adhesive, where
		\[\comma{\mathcal{N}}{\mathcal{M}}:=\{(h,k) \in \mor(\comma{L}{R}) \mid h\in \mathcal{N}, k\in \mathcal{M}\}\]
		\item for every object $X$
		the categories $\X/X$  and $X/\X$ are, respectively, (strictly) $\mathcal{M}/X$-adhesive and (strictly) $X/\mathcal{M}$-adhesive, where
		\[\mathcal{M}/X:=\{m\in \mor(\X/X) \mid m\in
		\mathcal{M}\} \hspace{25pt} X/\mathcal{M}:=\{m\in \mor(X/\X) \mid m\in \mathcal{M}\}\]
		\item for every small category $\Y$, the category $\X^\Y$ of
		functors $\Y\to \X$ is (strictly) $\mathcal{M}^{\Y}$-adhesive, where
		$\mathcal{M}^{\Y}:=\{\eta \in \mor(\X^\Y) \mid \eta_Y \in
		\mathcal{M} \text{ for every } Y\in \Y\}$;
		\item if $\Y$ is a full subcategory of $\X$ closed under pullbacks and $\mathcal{M}$-pushouts, then $\Y$ is (strictly) $\mathcal{N}$-adhesive for every class of arrows $\mathcal{N}$ of $\Y$ contained in $\mathcal{M}$ that is stable under pullbacks and pushouts, contains all isos, and is closed under composition and decomposition.
	\end{enumerate} 
\end{thm}

@@

\begin{exa}
	\label{ex:adhesive}
	$\cat{Set}$ is adhesive and, in fact, every topos is
	adhesive~\cite{lack2006toposes}. By the closure properties above, every presheaf $[\cat{X},\cat{Set}]$ is adhesive, thus the category
	$\cat{Graph} = [ E \rightrightarrows V, \cat{Set}]$ is adhesive
	where $E \rightrightarrows {V}$ is the two objects category with two
	morphisms $s,t \colon{E} \to {V}$. Similarly, various
	categories of hypergraphs can be shown to be adhesive, such as term
	graphs and hierarchical graphs~\cite{CastelnovoGM24}. Note that the category $\cat{sGraphs}$ of simple graphs, 
	i.e.~graphs without parallel edges, is
	$\reg{(\cat{sGraphs})}$-adhesive~\cite{BehrHK23} but not
	quasiadhesive.
\end{exa}

@@

We can state some useful properties of $\mathcal{M}$-adhesive category
(see, for instance, \cite[Thm.~4.26]{ehrig2006fundamentals} or \cite[Fact 2.6]{azzi2019essence}). A proof is provided in \Cref{regmono-proof}.
\begin{restatable}{prop}{rgmono}\label{prop:regmono}
	Let $\X$ be an $\mathcal{M}$-adhesive category. Then the following facts hold
	\begin{enumerate}
		\item every $\mathcal{M}$-pushout square is also a pullback;
		\item every arrow in $\mathcal{M}$ is a regular mono.
	\end{enumerate}
\end{restatable}
%  [Proof in ]

@@

\begin{restatable}{prop}{epic}\label{prop:reg_epi_coeq}
    Let $e\colon X \eto Y$ be a regular epi in a category $\X$ with a kernel pair $(K_e, \pi^1_e, \pi^2_e)$. Then $e$ is the coequalizer of $\pi^1_e$ and $\pi^2_e$.
\end{restatable}
%[Proof in \Cref{epic-proof}]

@@

		Moreover, the following facts hold
		\begin{enumerate}
			\item if $h$ is a mono then $k_h$ is a mono;
			\item if the leftmost square is a pullback then the central two are pullbacks;
			\item if $h$ is mono and the leftmost square is a pullback then the rightmost is a pullback.
		\end{enumerate} 
\end{restatable}
%[Proof in \Cref{kpp-proof}]

@@

\noindent
\begin{minipage}[l]{.6\linewidth}
\begin{restatable}{prop}{mpo}\label{lem:mpo}
	Let $\X$ be a strict $\mathcal{M}$-adhesive category with all pullbacks, and suppose that in the cube aside the top face is an $\mathcal{M}$-pushout and all the vertical faces are pullbacks. Then the right square is a pushout.
\end{restatable}
\end{minipage}\hfill 
\begin{minipage}[r]{.38\linewidth}\xymatrix@C=10pt@R=1pt{&A'\ar[dd]|\hole_(.65){a}\ar[rr]^{f'} \ar@{>->}[dl]_(.55){m'} && B' \ar[dd]^{b} \ar@{>->}[dl]_(.55){n'} & K_a\ar[rr]^{k_{f'}} \ar[dd]_{k_{m'}}&& K_b \ar[dd]^{k_{n'}} \\ C'  \ar[dd]_{c}\ar[rr]^(.7){g'} & & D' \ar[dd]_(.3){d}\\&A\ar[rr]|\hole^(.65){f} \ar[dl]_{m} && B \ar[dl]^{n}  & K_{c} \ar[rr]^{k_{g'}}&& K_d\\C \ar[rr]^{g} & & D }
	\end{minipage}

@@

\begin{cor}\label[corollary]{cor:monhyper}
	A morphism $(h,k)$ is a mono in $\hyp$ if and only if its components are injective functions.
\end{cor}

@@

\subsubsection{Two  subcategories of $\hyp$}

We are now going to introduce two classes of hypergraphs, and the relative subcategories, which will provide the base for our definitions of EGGs  (cfr.~\Cref{sec:sepegg,sec:lmegg}).
